Stability and Change in Personal Culture

### 2019-12-06 manuscript9.Rmd
minor edits

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-01.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave. This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change. Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development. However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items. As such, we designate a second term, _consistency_, to reflect this second dimension of change.

@@

_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate peripheral attitudes, such as views on specific policy questions [@boutylinevaisey; @goldbergstein]. Much work in political science has sought to understand whether people use ideological identity, policy positions, moral beliefs, or partisanship to adjudicate their position on other issues [@lakoff; @greenetal; @kinderkalmoe]. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification as the "core" belief in the political space, consistent with Converse's [-@converse] notion of social constraint. Baldassari and Gelman [-@baldassarigelman] find an increased alignment between partisan identification and policy positions, but little increase in the correlation between specific policy positions over time. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442]. This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate a highest level.

These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate a cascade of changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs. In our framework, beliefs and behaviors that exhibit stable dispositions are more likely to be “core” beliefs than those that exhibit active updating. Core beliefs should also be more consistently reported, since individuals should know where they stand on these core questions. Determining where one stands on peripheral issues requires greater cognitive effort and therefore are more prone to random perturbations, so should be reported less consistently.

@@

Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse]. This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these. However, we do spotlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population. In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, we predict active updating throughout the population, not just young people.

@@

First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo]. This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.

@@

At the same time, our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating in responses. We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model. If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the population average, and $\beta = 0$. If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.

@@

Finally, a number of approaches exist for evaluating theoretical process of belief change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization]. These tools are well designed to address the questions they set out to. However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.

@@

To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates. Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2. Under the settled disposition model, these responses would reflect either measurement error or a temporary shift at either wave 1 or wave 2, and we would expect that about 50 percent or 129 individuals would maintain the same response into wave 3. Only 147 of the 257 people (57 percent) who changed between waves 1 and 2 maintained the same response at wave 3. Therefore, we only have evidence that about 18 individuals or less than one percent of the sample showed evidence of persisting change. In other words, even for items that show strong evidence of active updating, the overall rate of attitude change in the population is likely small.

The right side of Figure \@ref(fig:phihist) shows that while the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period. That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position. This group includes many items about abortion, civil liberties, confidence in institutions, and views on race and gender.

@@

### Political Beliefs

@@

First, _are patterns of cultural change generally better described by the active updating model or the settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model. Around 40 percent of all items show no evidence for updating. For the items that do show evidence of active updating, the overall rate of change in the population is low. Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to where they were. What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago. Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is. But the average level of updating for most views appears to be low.

@@

The overall pattern of results, while diverse and at times hard to reconcile, does shed light on theoretical debates, to which we now turn.

@@

What are the broader implications of these findings, both for theories of cultural change and for empirical work in this area?

In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly, through the mechanism of cohort succession. Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled by early adulthood. The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time. In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating. Even for those items that did display strong evidence for active updating (e.g. "whites rich or poor" and "whites work hard") were still hard to predict wave after wave.

@@

For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change. Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to factors that encourage openness to change at different ages. Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes. Doing so will expand our understanding of the social and psychological factors that that give rise to stable or variable attitudes [@howekrosnick]. Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is a strong start [@eatonetal; @vissermirabile].

@@

We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning. We see too much inconsistency in responses to most questions for models of commitment bias to be applicable. Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p. 499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p. 500], or questions that “induce respondents to provide socially non-normative or stigmatized responses” [@warrenhalpernmanners: p. 501], tend to show low active updating and low consistency.

We cannot and would not want to rule out the possibility that panel conditioning is taking place in the GSS, and we believe it would be worthwhile to explore these same GSS panels for evidence of panel conditioning. However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.

>>>>

---
title: "Stability and Change in Personal Culture"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

@@

Neither model makes any assumptions about the magnitude of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave. This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change. Knowing that there is more wave-to-wave change in one item than another tells us little about how these two items relate to underlying theories of attitude development. However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items. As such, we designate a second term, _consistency_, to reflect this second dimension of change. _Consistency_ here simply refers to the degree that the attitude development process departs from pure randomness.

@@

_Core-periphery models:_ A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate peripheral attitudes, such as views on specific policy questions [@boutylinevaisey; @goldbergstein]. Much work in political science has sought to understand whether people use ideological identity, policy positions, moral beliefs, or partisanship to adjudicate their position on other issues [@lakoff; @greenetal; @kinderkalmoe]. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find support for ideological identification as the "core" belief in the political space, consistent with Converse's [-@converse] notion of social constraint. Baldassari and Gelman [-@baldassarigelman] find an increased alignment between partisan identification and policy positions, but little increase in the correlation between specific policy positions over time. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442]. This suggests that among items tapping issues of politics, most policy issues should show little active updating, but partisan affiliation should demonstrate the highest level.

These core-periphery models suggest that core beliefs should be relatively stable, that changes in core beliefs should necessitate changes in peripheral beliefs, and that people attempt to align their peripheral beliefs with their core beliefs. In our framework, beliefs and behaviors that exhibit stable dispositions are more likely to be “core” beliefs than those that exhibit active updating. Core beliefs should also be more consistently reported, since individuals should know where they stand on these core questions. Determining where one stands on peripheral issues requires greater cognitive effort and therefore are more prone to random perturbations, so should be reported less consistently.

@@

Theories that argue that individuals do not hold consistent opinions also posit that issues that receive significant media attention can be reported more consistently than other issues [@zaller; @converse]. This means a small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these. However, we do highlight one issue that saw significant change in elite opinion, which should lead to active updating in the general population. In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, we expect to find evidence of active updating throughout the population, not just among young people.

@@

First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ Previous work on cultural change using cross-sectional data has suggested that cohort effects are generally more important than period effects in explaining broad cultural change [@vaiseylizardo]. This implies that, in the repeated measures data on adults we will use here, we should find that the settled dispositions model performs better on most items because cohort formation should be (mostly) complete.

@@

At the same time, our estimates of $\beta$ provide a measure of the consistency of individuals' responses, contingent on the degree of active updating in responses. We can think of this parameter as analogous to an $R^2$ measure in a traditional linear model, capturing the total "predictiveness" of the model. If individuals pick a random response at each wave, the best predictor for a person at wave 3 will be the sample average, and $\beta = 0$. If there is little random fluctuation between waves, once the amount of active updating is accounted for,  $\beta$ will approach 1.

@@

Finally, a number of approaches exist for evaluating theoretical process of belief change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization]. These tools are well designed to address the questions they set out to answer. However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.

@@

To provide a concrete example, consider the GSS question that asks respondents whether they think it should be possible for a woman to receive a legal abortion if she became pregnant as a result of rape, to which individuals can respond either "yes" or "no." This produces a $\phi$ estimate of 0.62, above the 75th percentile of all $\phi$ estimates. Of the 2259 people who responded to the question in three waves, 257 changed between waves 1 and 2. Under the settled disposition model, these responses would reflect either measurement error or a temporary shift at either wave 1 or wave 2, and we would expect that about 50 percent or 129 individuals would maintain the same response into wave 3. Only 147 of the 257 people (57 percent) who changed between waves 1 and 2 maintained the same response at wave 3. Therefore, we only have evidence that about 18 individuals (less than one percent of the sample) showed evidence of persisting change. In other words, even for items that show strong evidence of active updating, the overall rate of attitude change in the population is likely small.

The right side of Figure \@ref(fig:phihist) shows that although the majority of items prefer the free parameter, 73 items (about 40 percent of the total) prefer the constraint, meaning these items show no evidence of active updating over this period. That is, although respondents might give different answers to these items in any particular wave because of measurement error or a transient change of opinion, they tend to revert to their previous position. This group includes many items about abortion, civil liberties, confidence in institutions, and views on race and gender.

@@

## Political Beliefs

@@

First, _to what extent are patterns of cultural change generally better described by an active updating model or a settled dispositions model?_ In general terms, consistent with previous research (e.g., Vaisey and Lizardo 2016), we see a greater degree of evidence in support of the settled dispositions model. Around 40 percent of all items show no evidence for updating. For the items that do show evidence of active updating, the overall rate of change in the population is likely low. Adult Americans are highly consistent in their attitudes from year to year, and those who appear to change tend to bounce back to where they were. What that means in practice is that knowing what a person said two years ago provides almost no better prediction of their current views than knowing what they said four years ago. Unfortunately, because of measurement error, we cannot be sure exactly how much updating there is. But the average level of updating for most views appears to be low.

@@

The overall pattern of results, while diverse and at times hard to reconcile, does shed light on several theoretical debates. We now consider these implications.

@@

What are the broader implications of these findings, both for theories of cultural change and for empirical work in this area? In the domain of cultural theory, our findings support the view that a great deal of cultural change happens slowly, through the mechanism of cohort succession. Most beliefs about gender roles, sexual morality, and abortion are, by and large, settled by early adulthood. The settled nature of these beliefs is often coupled with a high rate of consistency, suggesting that individuals truly hold these beliefs or at least have sufficient external support to consistently report them over time. In contrast, many views about race were so inconsistent that it would be difficult to call them either settled or updating. Even for those items that did display strong evidence for active updating (e.g. "whites rich or poor" and "whites work hard") were still hard to predict wave after wave.

@@

For some beliefs, such as whether aging parents should live with their adult children and whether divorce laws are too lenient, persistent change becomes more common with age, a pattern that is not accounted for in any major hypothesis about the relationship between age and attitude change. Rather than supporting a single theory linking age to attitude change, our results call for more work linking attitude content to social factors and institutional norms that encourage openness to change at different ages. Rather than assuming that "attitudes" in general are more or less likely to change at particular ages, we should explore the relationship between age and stability for a range of attitudes. Doing so will expand our understanding of the institutional and developmental factors that that give rise to stable or variable attitudes [@howekrosnick]. Recent work that has attempted to provide social explanations for (mostly political) attitude stability in middle age is an important step in that direction [@eatonetal; @vissermirabile].

@@

We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning. We see too much inconsistency in responses to most questions for models of commitment bias to be a major explanatory factor. Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p. 499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p. 500], or questions that “induce respondents to provide socially non-normative or stigmatized responses” [@warrenhalpernmanners: p. 501], tend to show low active updating and low consistency.

We cannot (and would not want to) rule out the possibility that panel conditioning is taking place in the GSS. We believe it is worthwhile to explore these same GSS panels for evidence of panel conditioning. However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.
